Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-04-17 manuscript/discards.qmd
update intro/figure captions

>>>>

The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g. @Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990]. Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

constrain palaeoclimatic conditions, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Wolfe1995; @Kvavcek2007; @Cao2019]. Instead, palaeoclimate signals (3)

### 2023-04-17 manuscript/main.qmd
update intro/figure captions

@@

@@

----

The early Eocene is the focus of a substantial body of palaeoclimate literature [e.g. @Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990]. Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

how ecosystems respond to climatic and environmental change. To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023]. Geochemical proxies such as oxygen isotopes (𝛅^18^O) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions. However, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

The recent increasing availability of large-scale compilations of deep-time climate proxy data provide an unprecedented opportunity to reconstruct palaeoclimatic conditions [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022]. However, two fundamental challenges must still be addressed to exploit the full potential of these data sets. First, disparate types of proxy data need to be integrated in order to generate unified palaeoclimatic reconstructions, making use of all available data. Second, the impact of incomplete preservation and uneven sampling of proxy records needs to be addressed [@Jones2022].

constrain palaeoclimatic conditions, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Wolfe1995; @Kvavcek2007; @Cao2019]. Instead, palaeoclimate signals (3)

Even when using the full range of available records, proxy-based climate reconstructions beyond the Holocene are inevitably patchy [e.g. @Hollis2019], and global temperature reconstructions based on those records alone are prone to bias [@Jones2022]. Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps. A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for a given sample size (5, 10, 20, and 50). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse (flat), greenhouse (quadratic), icehouse (linear), extreme icehouse (steep). The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs. The blue shading depicts the 90%, 95%, and 99% credible intervals. Each gradient is depicted in absolute latitude.
:::

@@

Estimates of the present-day latitudinal temperature gradient using the Bayesian hierarchical model. (A) The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperature. Grey points depict the individual cell values of the Bio-ORACLE grid. (B) The present-day empirical (black line) latitudinal temperature gradient and that estimated by the Bayesian hierarchical model (purple line) when limiting cell values by the palaeogeographic distribution of Eocene samples. To do so, present-day temperature values were extracted based on the palaeolatitudes of Eocene samples with randomly generated longitudes. This process was repeated 100 times, randomly choosing longitude for each sample in each repetition to capture the potential longitudinal variability in sea surface temperature. Purple points depict the median sea surface temperature value for each sample and the purple bars illustrate the 95% credible intervals.
:::

@@

Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model. The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient. The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients. The yellow points depict the palaeolatitudinal distribution of geochemical data (e.g. \$\\delta\$^18^O) and their respective point estimate temperature values. The green squares depict the palaeolatitudinal distribution of geoecological data (i.e. coral reefs and mangroves assemblages) and their mean temperature values sampled from their respective normal distributions.
:::

@@

@@

@@

>>>>

To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023]. Geochemical proxies such as oxygen isotopes ($\delta^18^O$) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions. Nevertheless, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps. A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (sample size for the EECO). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs. The blue shading depicts the 90%, 95%, and 99% credible intervals. Each gradient is depicted in absolute latitude.
:::

@@

Estimate of the present-day latitudinal temperature gradient using the Bayesian hierarchical model. The present-day empirical (black line) latitudinal temperature gradient (mean sea surface temperature) and that estimated by the Bayesian hierarchical model (green line) when using all cell values from the Bio-ORACLE grid of mean sea surface temperature. Grey points depict the individual cell values of the Bio-ORACLE grid.
:::

@@

Estimates of the present-day and early Eocene climatic optimum latitudinal temperature gradient using the Bayesian hierarchical model. The estimated present-day (green line) and early Eocene climatic optimum (purple line) latitudinal temperature gradient. The respective ribbons (green and purple shading) depict the 95% credible intervals of the estimated gradients. Points within the plot depict the palaeolatitudinal distribution of geochemical (e.g. $\delta^18^O$) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted by their mean temperature values sampled from their respective normal distributions.
:::

@@

The authors are grateful to all those who have enabled this work by collecting, measuring, collating and screening geochemical and fossil data. The contribution of K.E. was supported by... The contribution of L.A.J. was supported by a Juan de la Cierva-formación 2021 fellowship (FJC2021-046695-I/MCIN/AEI/10.13039/501100011033) from the European Union "NextGenerationEU"/PRTR.

@@

Both authors designed the study and carried out data preparation. K.E. programmed the model and conducted the analyses. L.A.J. generated the figures. Both authors contributed to the writing of the manuscript.

# Competing Interests

The authors declare that they have no conflicts of interest.

@@

The data and code used to produce the results of this study are available via GitHub (<https://github.com/KEichenseer/PalaeoClimateGradient>) and the linked Zenodo repository (XXX).
